Retrofitting Effect Handlers onto OCaml

### 2020-11-12 retro-concurrency.tex
iintro edits

Effect Handlers~\cite{Plotkin09} provide a modular foundation for user-defined
effects. The key idea is to separate the definition of the effectful operations
from their interpretations, which are defined by \emph{handlers} of the
effects. For example,
%
\begin{lstlisting}
effect Read : in_channel -> string
\end{lstlisting}
%
declares an \emph{effect} |Read|, which is parameterised with an input channel
of type |in_channel|, which when \emph{performed} returns a |string| value. A
computation can perform the |Read| effect without knowing how the |Read| effect
is implemented. This computation may be enclosed by different handlers that
interpret |Read| differently. For example, |Read| may be implemented by
performing a blocking read on the input channel or performing the read
asynchronously by offloading it to an event loop such as |libuv|. Thanks to the
separation of effectful operations from their interpretations, effect handlers
enable new approaches to modular programming.

@@

What makes effect handlers quite powerful is that they provide a structured
form of delimited continuations~\cite{Danvy90}, which enables advanced
non-local control-flow mechanisms such as resumable exceptions, coroutines,
generators and asynchronous I/O to be composably expressed. As an example,
consider the task of handling division-by-zero case in a calculator
application. Different calculators handle division-by-zero differently. In the
iOS 13 Calculator App (|iOS|), division-by-zero returns |Error|. In the Google
search calculator App (|gsearch|) |0/0| returns to |Error|, and for any other
|n| not equal to |0|, |n/0| returns |Infinity| with the same sign as |n|. How
can we capture these different behaviours without changing the evaluator?

Common Lisp Conditions~\cite{Steele90} are an elegant way to solve this
problem. Unlike exceptions where the library decides whether an error is fatal
or not, conditions allow the client of the library to make that decision. If
the error is non-fatal, Conditions provide a \emph{restart} mechanism to resume
execution from the point where the condition was signalled. On
division-by-zero, the calculator signals the handler and lets the handler
decide the course of action based on the desired semantics. We can implement
this mechanism using effect handlers as shown below:

@@

On line 4, we declare an effect |DivBy0| that is parameterised by an integer,
which when performed returns the result of division-by-zero. If the denominator
is 0, then we perform |DivBy0| effect with the numerator value (line 12) so that
the context may decide the return value. We can get the |iOS| semantics by
wrapping the evaluator with the following handler:
\begin{lstlisting}
let ios e = match eval e with
| v -> v
| effect (DivBy0 _) k -> continue k Error
\end{lstlisting}
If the evaluator returns a value, we simply return that. If the evaluator
performs the |DivBy0| effect, we catch the effect in the effect case along with
the continuation |k| of the corresponding |perform| \emph{delimited} by this
handler. In the case of |iOS|, we |continue| the suspended computation with
|Error| value. Hence, the computation |ios(1/0)| returns |Error|. On the
other hand, we can get the |gsearch| semantics by enclosing the evaluator in the
following handler:
\begin{lstlisting}
let gsearch e = match eval e with
| v -> v
| effect (DivBy0 v) k ->
		if v = 0 then continue k Error
		else if v > 0 then continue k (Inf P)
		else continue k (Inf N)
\end{lstlisting}
The computation |gsearch(1/0)| returns positive infinity (|Inf P|).

>>>>

Effect handlers~\cite{Plotkin09} provide a modular foundation for user-defined
effects. The key idea is to separate the definition of the effectful operations
from their interpretations, which are defined by \emph{handlers} of the
effects. For example,
%
\begin{lstlisting}
effect Read : in_channel -> string
\end{lstlisting}
%
declares an \emph{effect} |Read|, which is parameterised with an input channel
of type |in_channel|, which when \emph{performed} returns a |string| value. A
computation can perform the |Read| effect without knowing how the |Read| effect
is implemented. This computation may be enclosed by different handlers that
interpret |Read| differently. For example, |Read| may be implemented by
performing a blocking read on the input channel or performing the read
asynchronously by offloading it to an event loop such as |libuv|. Thanks to the
separation of effectful operations from their interpretations, effect handlers
enable new approaches to modular programming.

@@

Effect handlers provide a structured
form of delimited continuations~\cite{Danvy90} that enables
non-local control-flow mechanisms such as resumable exceptions, coroutines,
generators and asynchronous I/O to be composably expressed. For example,
consider the task of handling division-by-zero in a calculator
application. Different calculators handle division-by-zero differently. In the
iOS 13 Calculator (|iOS|), |0/0| returns |Error|. In the Google
search calculator (|gsearch|) |0/0| returns |Error|, but for
|n!=0| the |n/0| returns |Infinity| with the same sign as |n|.
How do we capture all these different behaviours without changing the evaluator?

Common Lisp Conditions~\cite{Steele90} are an elegant way to solve this
problem. Unlike exceptions where the library decides whether an error is fatal
or not, conditions allow the client of the library to make that decision.
Conditions provide a \emph{restart} mechanism for non-fatal errors to resume
execution from the point where the condition was signalled. On
division-by-zero, the calculator signals the handler and lets the handler
decide the course of action based on the desired semantics. We now implement
this using effect handlers in OCaml syntax:

@@

Line 4 declares an effect |DivBy0| that is parameterised by an integer,
which when performed returns the result of division-by-zero. If the denominator
is 0, then we perform the |DivBy0| effect with the numerator value (line 12) so that
the context may decide the return value. We obtain the |iOS| semantics by
wrapping the evaluator with this handler:
\begin{lstlisting}
let ios e = match eval e with
| v -> v
| effect (DivBy0 _) k -> continue k Error
\end{lstlisting}
If the evaluator returns a value, we simply return that. If the evaluator
performs the |DivBy0| effect, we catch the effect in the pattern match along with
the continuation |k| of the corresponding |perform| \emph{delimited} by this
handler. For |iOS|, we |continue| the suspended computation with an
|Error| value and the computation |ios(1/0)| returns |Error|.
We can get the |gsearch| semantics by enclosing the evaluator in this handler:
\begin{lstlisting}
let gsearch e = match eval e with
| v -> v
| effect (DivBy0 v) k ->
		if v = 0 then continue k Error
		else if v > 0 then continue k (Inf P)
		else continue k (Inf N)
\end{lstlisting}
Executing |gsearch(1/0)| returns positive infinity (|Inf P|).
